---
title: "Instrumental variables"
author: "Mark Deming"
date: "3/19/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Basic intuition
We want to estimate the impact of $X$ on $Y$. But some third variable may influence both $X$ and $Y$. Or, there may be reverse causation (aka, **simultaneity**): $Y$ may influence $X$.

We can control for some of these third variables using multivariate regression. But we cannot control for everything. We may therefore turn to IV regression. In IV regression, we identify a third variable $Z$ that is related to $Y$ **onlty through $X$**>


## Assumptions
To work, IV regression requires 2 assumptions:

- Relevance
- Validity

### Relevance
This one is easy: does the IV ($Z$) predict $X$? We can check for relevance by calculating the covariance of $X$ and $Z$.

### Validity
This one is harder: is the IV related to $Y$ *only* through $X$? This assumption is commonly known as the **exclusion restriction** in political science. Satisfying the exclusion restriction is, above all, about **logic**. 

Consider Huntington-Klein's example about Acemoglu and Robinson's 2007 article. The authors examine the impact of populations' health on long-run economic growth. They argue that the introduction of a new medical technology such as a vaccine is a good instrument for medical health.

Why? The authors argue that (1) the introduction of a vaccine is related to health (relevance). More importantly, (2) the introduction of a vaccine should be related to long-run economic growth **only through improvements in health**. This is a logical argument. What do you think about it?

Bloom, Canning, and Frank (2014) did not buy the argument. They pointed out that (again, logically), vaccinations' impact on health is itself shaped by economic development, since higher development weakens the impact of vaccinations on health in the long run. Put differently, vaccinations are not random, or **exogenous**.

The **key point** of the above example is that meeting the exclusion restriction is about making an argument on logical grounds. Not everyone will buy the logic.

## Procedure

We run IV regression in 2 stages. Imagine that we have:

- outcome variable $y$
- treatment variable $x$
- control variables $w1$, $w2$, etc.
- instrumental variable $y$

### Stage 1

In R: 

```{r eval = FALSE}
first_stage_model <- lm(x ~ z + w1 + w2..., data)
```

Then, we predict the values of $x$ using the results of that regression:

```{r eval = FALSE}
predict(first_stage_model)
```

### Stage 2

We regress $y$ on the predicted values ($x_hat$) that we calculated above, along with controls:

```{r eval = FALSE}
second_stage_model <- lm(y ~ x_hat + w1 + w2..., data)
```

### Using `ivreg`

In practice, we can use the `ivreg` function from the `AER` package:

```{r eval = FALSE}
library(ivreg)
ivreg_model <- ivreg(y ~ x1 + w1 + w2... | z, data)
```

Or, we can use `feols` from the `fixest` package (as in Huntington-Klein):

```{r eval = FALSE}
library(fixest)
feols_model <- feols(y ~ x1 + w1 + w2... | z, data)
```

## Example

The example below is a simplified version of an example from Huntington-Klein's, *The Effect*. I run the example 3 ways:

- "by hand"
- `ivreg`
- `feols`

```{r}
# load packages
library(tidyverse) 
library(AER)
library(modelsummary)
library(fixest)
library(causaldata)

# load data
data <- causaldata::social_insure
```

**By hand**
```{r}
# First stage: x ~ z + w1 + w2...
stage1 <- lm(pre_takeup_rate ~
             default +          
             male + 
             age + 
             agpop + 
             ricearea_2010 + 
             literacy + 
             intensive + 
             risk_averse + 
             disaster_prob, 
             data = data)
coeftest(stage1, vcov = vcovHC, type = "HC1")

# Grab fitted values
data$x_hat <- predict(stage1, newdata = data)

# Second stage: y ~ x_hat + w1 + w2...
stage2 <- lm(takeup_survey ~ x_hat + 
             male + 
             age + 
             agpop + 
             ricearea_2010 + 
             literacy + 
             intensive +
             risk_averse + 
             disaster_prob, 
             data = data)
coeftest(stage2,  vcov = vcovHC, type = "HC1")
```

- With `ivreg`
```{r}
ivreg_model <- ivreg(takeup_survey ~ pre_takeup_rate + male + age + agpop + ricearea_2010 + 
                     literacy + intensive + risk_averse + disaster_prob |
                     default + male + age + agpop + ricearea_2010 + literacy + 
                     intensive + risk_averse + disaster_prob, 
                     data = data)
coeftest(ivreg_model, vcov = vcovHC, type = "HC1")
```

- With `feols`
```{r}
feols_model <- feols(takeup_survey ~ male + age + agpop + ricearea_2010 + literacy + 
                     intensive + risk_averse + disaster_prob | pre_takeup_rate ~ default,
                     data = data)
summary(feols_model, vcov = "HC1")
```