---
title: "Assignment 2 - Solutions"
author: "Mark Deming"
date: "Due 2/26"
output: pdf_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

## Overview

You will do a complete replication of Table 4 from Koch and Nicholson's article (2016), "Death and Turnout". This is part 2 of 2.

You will require the four datasets that you used in Assignment 1. All datasets are at Deming's GitHub page: [HERE](https://github.com/jmdeming/independent_study):

-   `bes05_short.dta`
-   `bes10_short.dta`
-   `ukregion_cas.tab`
-   `0501districtdata.tab`

Throughout, you should use `dplyr` functions and syntax whenever possible.

### Get started

**1.** Load the packages below. You may need to install them first.

-   `tidyverse` (contains `dplyr` and `ggplot2`)
-   `readr` (for importing `.tab` formatted data)
-   `haven` (for importing `.dta` formatted data)
-   `here` (recommended but not required. You might read about how `here()` works.)
-   `estimatr` (for robust standard errors)
-   `modelsummary` (for nice regression tables)
-   `plm` (for panel data analysis)

**2.** Import the four datasets above in the `setup` chunk.

**3.** In the chunk below, clean, append, and merge the datasets as you did in Assignment 1 (up through the "Explore" section). See if you can clean `bes05` and `bes10` in two long strings of piped code.

```{r}

# 1. Clean bes05 in a single string of piped code

# 2. Clean bes10 in a single string of piped code

# 3. Append bes05 and bes10

# 4. Merge the appended bes dataset with casualties

# 5. Merge the bes-casualties dataset with the district data. Then, create the female, 
# low_attention, married, and partstrength variables. Do this in a single string
# of piped code.

```

# Ordinary least squares model

**4.** Use `lm_robust()` to write a regression model that approximates Table 4 from Koch and Nicholson (2016). Do not include region dummies in the model.

**5.** Use `modelsummary()` to display your results above in a nice regression table. Add informative coefficient labels following Table 4.

**6.** Write a short paragraph of interpretation of the regression results above. Focus on the main coefficients: `region_cas`, `low_attention`, and the interaction term `low_attention:region_cas`. (What does the interaction term denote?)

## Random effects model

The `bes` data are panel data: the same respondents are surveyed in 2005 and 2010. We can use a random effects model to account for the panel structure.

**7.** Create a new dataset that is a duplicate the Koch and Nicholson data above (2016). Designate the dataset as panel data using an appropriate function from the `plm` package. Use the `besid` and `year` variables as the index.

**8.** Use `plm()` to write a random effects model that closely approximates Table 4 from Koch and Nicholson (2016). You should include region dummies in the model.

**9.** Write a short paragraph of interpretation of the regression results above. In addition to interpreting the main coefficients, write 2-3 sentences about the random effects model. What does the random effects model add to the OLS model?

# Wrap Up

When you finish:

-   Knit this RMD to PDF.
-   Review the RMD for completeness, accuracy, and neatness.
-   Submit the RMD and PDF.
