---
title: "Assignment 3"
subtitle: "Solutions"
author: "Mark Deming"
date: "3/02/2025"
output:
  pdf_document
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{labelformat=empty}
  - \captionsetup[table]{labelformat=empty}
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load packages

# Read in data=
```

# Overview
In this assignment, you will replicate portions of Card & Krueger's analysis of minimum wages and employment (1994). The assignment culminates in a complete replication of Table 4, in which the authors run a series of regressions using a differences-in-differences design (DiD).

# Getting started

**1.** Download the Card & Krueger (1994) data from the course Github site. You should also download the corresponding description of variables.

**2.** Start by loading the following packages in the `setup` chunk at the top of this RMD:

- tidyverse
- readr
- here
- modelsummary

**3.** Read in the Card & Krueger (1994) data. The data are in .csv format. Take some time to explore the data:

- structure (dimensions, cross-sectional vs. panel, etc.)
- variables (names, types, etc.)

# Cleaning & Transforming

**4.** The authors create a new variable, full-time equivalent employment (`fte`). They describe the variable on page 775. Follow the authors' definition to create 2 new variables: `fte1` and `fte2`. The variables should measure full-time employment for waves 1 and 2, respectively.

```{r}

```

**5.** The values of the variable `chain` denotes different restaurant chains. Use this variable to create 4 new dummy variables, one for each restaurant chain: e.g., `bk` = 1 if a restaurant is a Burger King and 0 otherwise.

```{r}

```

# Table 2 

**6.** Let's ensure that the data match the authors'. To do so, calculate and print the proportions shown in 1a-e, 2a, and 3a of Table 2 on page 776. It is *not* necessary to create a nice table.

```{r}

```

# Figure 1 

**7.** Use `ggplot` to replicate Figure 1, panel 1 (February 1992). **Hint:** Before creating the plot, you will need to bin `wage_st` into discrete categories. The relevant function is `cut()`.

```{r}

```

**8.** Use `ggplot` to replicate Figure 1, panel 2 (November 1992). As above, you will need to bin `wage_st2` into discrete categories.

```{r}

```

# Table 4 

**9.** You will now replicate the authors' DiD design in Table 4. To get started, create a new variable named `gap`. Use the variable definition given by the authors on page 779. **Hint:** You will need to use 2 logical conditions inside `ifelse()`.

```{r}

```

**10.** Filter the dataset to include only (1) rows with complete data for `fte`, `fte2`, `wage_st`, and `wage_st2` OR (2) rows for restaurants that closed in wave 2.

```{r}

```

**11.** Replicate Table 4, models I through V (p. 780). Specifically, write 5 regression models. Compare the model output against Table 4. It should be the same for models I-IV and very similar for Model V.

```{r}

```

**12.** Use `modelsummary()` to print a nice regression table of models I-V above.

```{r}

```

**13.** From a technical perspective, DiD is simple. Why is it a credible design despite its simplicity? Write your answer in a few sentences below.
