---
title: "Assignment 3"
subtitle: "Solutions"
author: "Mark Deming"
date: "3/02/2025"
output:
  pdf_document
header-includes:
  - \usepackage{caption}
  - \captionsetup[figure]{labelformat=empty}
  - \captionsetup[table]{labelformat=empty}
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load packages
library(tidyverse)
library(readr)
library(here)
library(modelsummary)
library(kableExtra)

# Read in data
ck <- read_csv(here("data","original","card-krueger_1994.csv"))
```

# Overview
In this assignment, you will replicate portions of Card & Krueger's analysis of minimum wages and employment (1994). The assignment culminates in a complete replication of Table 4, in which the authors run a series of regressions using a differences-in-differences design (DiD).

# Getting started

**1.** Download the Card & Krueger (1994) data from the course Github site. You should also download the corresponding description of variables.

**2.** Start by loading the following packages in the `setup` chunk at the top of this RMD:

- tidyverse
- readr
- here
- modelsummary

**3.** Read in the Card & Krueger (1994) data. The data are in .csv format. Take some time to explore the data:

- structure (dimensions, cross-sectional vs. panel, etc.)
- variables (names, types, etc.)


# Cleaning & Transforming

**4.** The authors create a new variable, full-time equivalent employment (`fte`). They describe the variable on page 775. Follow the authors' definition to create 2 new variables: `fte1` and `fte2`. The variables should measure full-time employment for waves 1 and 2, respectively.

```{r}
ck <- ck %>%
  mutate(fte1 = empft + nmgrs + emppt*0.5,
         fte2 = empft2 + nmgrs2 + emppt2*0.5)
```

**5.** The values of the variable `chain` denotes different restaurant chains. Use this variable to create 4 new dummy variables, one for each restaurant chain: e.g., `bk` = 1 if a restaurant is a Burger King and 0 otherwise.

```{r}
ck <- ck %>%
  mutate(bk = case_when(chain == 1 ~ 1, chain != 1 ~ 0),
         kfc = case_when(chain == 2 ~ 1, chain != 2 ~ 0),
         roys = case_when(chain == 3 ~ 1, chain != 3 ~ 0),
         wendys = case_when(chain == 4 ~ 1, chain != 4 ~ 0))
```


# Table 2 

**6.** Let's ensure that the data match the authors'. To do so, calculate and print the proportions shown in 1a-e, 2a, and 3a of Table 2 on page 776. It is *not* necessary to create a nice table.

```{r}
# Section 1: Chains
(section1 <- ck %>%
  select(state, bk, kfc, roys, wendys, co_owned) %>%
  group_by(state) %>%
  summarize_all(mean) %>%
  
  # The code above is sufficient for the assignment.
  # The code below refines the dataframe. It is
  # included for reference.
  
  pivot_longer(cols = -state,
               names_to = "chain", 
               values_to = "value") %>%
  pivot_wider(id_cols = "chain",
              names_from = "state",
              values_from = "value"))

# Sections 2 and 3: FTE
(section2 <- ck %>%
  select(state, fte1, fte2) %>%
  group_by(state) %>%
  summarize_all(mean, na.rm = TRUE) %>%

  # The code above is sufficient for the assignment.
  # The code below refines the dataframe. It is 
  # included for reference.

  pivot_longer(cols = -state,
               names_to = "variable",
               values_to = "value") %>%
  pivot_wider(id_cols = "variable",
              names_from = "state",
              values_from = "value"))
```


# Figure 1 

**7.** Use `ggplot` to replicate Figure 1, panel 1 (February 1992). **Hint:** Before creating the plot, you will need to bin `wage_st` into discrete categories. The relevant function is `cut()`.

```{r}
ck <- ck %>%
  mutate(cuts = cut(wage_st, breaks = seq(4.20, 5.6, by = .10),
                             labels = seq(4.25, 5.6, by = .10)))
```

```{r}
# Basic Feb. 1992 plot
p1 <- ggplot(ck,              
             aes(x = cuts,                                   # Plot cuts, not wage
                 group = fct_rev(factor(state)),             # PA should be first
                 fill = fct_rev(factor(state)))) +           # PA should be first
  geom_bar(aes(y = after_stat(prop)),                        # Proportions along y-axis
           position = position_dodge(preserve = "single"))   # Maintain uniform bar width when count = 0
```

```{r}
# This code refines the basic plot above. It is not necessary to
# go this far with your own code. I provide the code for reference.

p1 + scale_fill_manual(values = c("black","gray70"),              # Fill color     
                       labels = c("Pennsylvania","New Jersey")) + # Legend labels 
  scale_y_continuous(limits = c(0, .35),                          # Y-axis limits
                     breaks = seq(0, .35, .05),                   # Y-axis breaks
                     labels = scales::percent,                    # Display Y-axis labels as percent
                     expand = c(0, 0)) +                          # Remove white space
  labs(title = "February 1992\n",                                 # Plot title
       x = "\nWage Range",                                        # X-axis title
       y = "Percent of Stores\n",                                 # Y-axis title
       fill = "State") +                                          # Legend title
  theme_minimal() +                                               # Add "minimal" theme
  theme(panel.grid.major.x = element_blank(),                     # Remove x-axis major grid lines
        panel.grid.major.y = element_line(linetype = 3,           # Adjust y-axis major grid lines
                                          color = "gray90"),      
        panel.grid.minor = element_blank(),                       # Remove all minor grid lines
        plot.title.position = "plot")                             # Push plot title all the way to left
```

**8.** Use `ggplot` to replicate Figure 1, panel 2 (November 1992). As above, you will need to bin `wage_st2` into discrete categories.

```{r}
ck <- ck %>%
  mutate(cuts2 = cut(wage_st2, breaks = seq(4.20, 5.6, by = .10),
                               labels = seq(4.25, 5.6, by = .10)))
```

```{r}
# Basic Nov. 1992 plot
p2 <- ggplot(ck,
             aes(x = cuts2,
                 group = fct_rev(factor(state)),
                 fill = fct_rev(factor(state)))) +
  geom_bar(aes(y = after_stat(prop)),
           position = position_dodge(preserve = "single"))
```

```{r}
# This code refines the basic plot above. It is not necessary to
# go this far with your own code. I provide the code for reference. 

p2 + scale_fill_manual(values = c("black","gray70"),              # Fill color     
                       labels = c("Pennsylvania","New Jersey")) + # Legend labels 
  scale_y_continuous(limits = c(0, .95),                          # Y-axis limits
                     breaks = seq(0, .95, .05),                   # Y-axis breaks
                     labels = scales::percent,                    # Display Y-axis labels as percent
                     expand = c(0, 0)) +                          # Remove white space
  labs(title = "November 1992\n",                                 # Plot title
       x = "\nWage Range",                                        # X-axis title
       y = "Percent of Stores\n",                                 # Y-axis title
       fill = "State") +                                          # Legend title
  theme_minimal() +                                               # Add "minimal" theme
  theme(panel.grid.major.x = element_blank(),                     # Remove x-axis major grid lines
        panel.grid.major.y = element_line(linetype = 3,           # Adjust y-axis major grid lines
                                          color = "gray90"),      
        panel.grid.minor = element_blank(),                       # Remove all minor grid lines
        plot.title.position = "plot")                             # Push plot title all the way to left
```

# Table 4 

**9.** You will now replicate the authors' DiD design in Table 4. To get started, create a new variable named `gap`. Use the variable definition given by the authors on page 779. **Hint:** You will need to use 2 logical conditions inside `ifelse()`.

```{r}
# Create new "gap" variable
ck <- ck %>%
  mutate(gap=ifelse(state ==1 & wage_st <= 5.05,((5.05 - wage_st) / wage_st), 0))
```

**10.** Filter the dataset to include only (1) rows with complete data for `fte`, `fte2`, `wage_st`, and `wage_st2` OR (2) rows for restaurants that closed in wave 2.

```{r}
# Select complete observations
ck <- ck %>%
  filter(complete.cases(fte1, fte2)) %>% 
  filter(complete.cases(wage_st, wage_st2) | status2 == 3)
```

**11.** Replicate Table 4, models I through V (p. 780). Specifically, write 5 regression models. Compare the model output against Table 4. It should be the same for models I-IV and very similar for Model V.

```{r}
# Model 1
mod1 <- lm(fte2 - fte1 ~ state, data = ck)

# Model 2
mod2 <- lm(fte2 - fte1 ~ state + I(bk) + I(kfc) + I(roys) + I(wendys) + I(co_owned), data = ck)

# Model 3
mod3 <- lm(fte2 - fte1 ~ gap, data = ck)

# Model 4
mod4 <- lm(fte2 - fte1 ~ gap + I(bk) + I(kfc) + I(roys) + I(wendys) + I(co_owned), data = ck)

# Model 5
mod5 <- lm(fte2 - fte1 ~ gap + I(bk) + I(kfc) + I(roys) + I(wendys) + I(co_owned) + 
                          I(centralj) + I(southj) + I(pa1) + I(pa2), data = ck)
```

**12.** Use `modelsummary()` to print a nice regression table of models I-V above.

```{r}
models <- list(mod1, mod2, mod3, mod4, mod5)

rows <- tibble::tribble(~IV, ~i, ~ii, ~iii, ~iv, ~v,
                        "Controls for chain and ownership", "no", "yes", "no", "yes", "yes",
                        "Controls for region", "no", "no", "no", "no", "yes")
attr(rows, "position") <- c(5, 6)

(table4 <- modelsummary(models,
                       stars = T,
                       title = "Reduced-form models for change in employment",
                       coef_map = c("state" = "New Jersey dummy",
                                    "gap" = "Initial wage gap"),
                       gof_omit = "DF|Deviance|R2|AIC|BIC|Log.Lik|F",
                       add_rows = rows,
                       ouput = "markdown"))
```

**13.** From a technical perspective, DiD is simple. Why is it a credible design despite its simplicity? Write your answer in a few sentences below.
