---
title: "Assignment 1"
author: "YOUR NAME GOES HERE"
date: 'Due 2/10'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages here
require(tidyverse)    # For data manipulation
require(readr)        # For reading in data
require(haven)        # For reading in Stata data
require(here)         # For setting working directory

# Load data here
bes05 <- read_dta(here("datasets","koch-nicholson_2016","bes05_short.dta"))
bes10 <- read_dta(here("datasets","koch-nicholson_2016","bes10_short.dta"))
casualties <- read_delim(here("datasets","koch-nicholson_2016","ukregion_cas.tab"), delim = "\t")
districts <- read_delim(here("datasets","koch-nicholson_2016","0501districtdata.tab"), delim = "\t")

```

## Overview

You will load, clean, manipulate, and explore data from Koch and Nicholson's article (2016), "Death and Turnout". This is part 1 of 2 of a complete replication of Table 4 from the article.

There are four relevant datasets. All datasets are at Deming's GitHub page under "independent_study/datasets/koch-nicholson_2016". [HERE](https://github.com/jmdeming/independent_study) is a link to Deming's GitHub page.

- `bes05_short.dta`
- `bes10_short.dta`
- `ukregion_cas.tab`
- `0501districtdata.tab`

Throughout, you should use `dplyr` functions and syntax whenever possible. 


### Get started

**1.** Load the following packages in the `setup` chunk. (You may need to install some of them first):

- `tidyverse` (contains `dplyr` and `ggplot2`)
- `readr`     (for importing `.tab` formatted data)
- `haven`     (for importing `.dta` formatted data)
- `here`      (recommended but not required. You might read about how `here()` works.)

**2.** Import the four datasets above in the `setup` chunk.

**3.** Take some time to get to know the four datasets before moving on. Examine their dimensions, variable names, etc. You may also wish to examine the `koch-nicholson_codebook` at GitHub, which contains variables descriptions for `bes05` and `bes10`.


## Clean and Manipulate

### 2005 BES data

**4.** Examine the variables in `bes05`. They are not very informative. The `koch-nicholson_codebook` at GitHub contains more informative names. Rename the variables according to codebook.

```{r}

# Rename variables
bes05 <- bes05 %>%
  rename(region = pre_q1,
         labor_iraq = pre_q13,
         conserve_iraq = pre_q23,
         partyid = pre_q29,
         party_strength = pre_q33,
         likelyvote = pre_q34,
         blair_competent = pre_q50,
         executive_approval = pre_q68,
         gov_party_approve = pre_q84,
         perception_economy = pre_q92,
         attention = pre_q141,
         birthyr = pre_q148,
         education = pre_q156,
         income = pre_q163,
         race = pre_q174,
         gender = pre_q180,
         marital_status = pre_q158,
         british_iraq = pre_q128,
         weights = pre_w8)

```

**5.** Some of the variables' current values actually denote missing data: "no response", "don't know", etc. Examine the codebook for the variables below. Recode them so that missing values are denoted as NA:

- `party_strength`
- `labor_iraq`
- `perception_economy`
- `likelyvote`

```{r}

# Recode missing values
bes05 <- bes05 %>%
  mutate(party_strength = if_else(party_strength == 4, NA_real_, party_strength),
         labor_iraq = if_else(labor_iraq == 6, NA_real_, labor_iraq),
         perception_economy = ifelse(perception_economy == 6, NA_real_, perception_economy),
         likelyvote = ifelse(likelyvote == 12, NA_real_, likelyvote))

```

**6.** Rename `executive_approval` to `pmtherm`. Also rename `labor_iraq` to `pmwar`.

```{r}

# Rename EXECUTIVE_APPROVAL and LABOR_IRAQ
bes05 <- bes05 %>%
  rename(pmtherm = executive_approval,
         pmwar = labor_iraq)
```

**7.** Create two new variables. Add them to `bes05`. Here are the variable definitions:

- `year`: equals 2005
- `age`: equals 2005 minus individuals' birth year

```{r}

# Add AGE and YEAR variables
bes05 <- bes05 %>%
  mutate(year = 2005,
         age = 2005 - birthyr)

```

### 2010 BES Data

**8.** Examine the variables in `bes10`. They are not very informative. The `koch-nicholson_codebook` at GitHub contains more informative names. Rename the variables according to codebook.

```{r}

# Rename variables
bes10 <- bes10 %>%
  rename(region = aaq1, 
         labor_afghan = aaq13, 
         conserve_afghan = aaq22, 
         partyid = aaq28, 
         party_strength = aaq32, 
         likelyvote = aaq33, 
         brown_competent = aaq81, 
         executive_approval = aaq52, 
         gov_party_approve = aaq63, 
         perception_economy = aaq87, 
         attention = aaq131, 
         birthyr = aaq151, 
         education = aaq159, 
         income = aaq166, 
         race = aaq177, 
         gender = aaq186, 
         marital_status = aaq161, 
         british_afghan = aaq116, 
         weights = w8_f)

```

**9.** Some of the variables' current values actually denote missing data: "no response", "don't know", etc. Examine the codebook for the variables below. Recode them so that missing values are denoted as NA:

- `party_strength`
- `labor_afghan`
- `perception_economy`
- `likelyvote`
- `income`

```{r}

# Recode missing values
bes10 <- bes10 %>%
  mutate(party_strength = if_else(party_strength == 4, NA_real_, party_strength),
         labor_afghan = ifelse(labor_afghan == 6, NA_real_, labor_afghan),
         perception_economy = ifelse(perception_economy == 6, NA_real_, perception_economy),
         likelyvote = ifelse(likelyvote == 12, NA_real_, likelyvote),
         income = ifelse(income == 17, NA_real_, income))

```

**10.** Rename `executive_approval` to `pmtherm`. Also rename `labor_iraq` to `pmwar`.

```{r}

# Rename EXECUTIVE_APPROVAL and LABOR_AFGHAN
bes10 <- bes10 %>%
  rename(pmtherm = executive_approval,
         pmwar = labor_afghan)

```

**11.** Create two new variables. Add them to `bes10`. Here are the variable definitions:

- `year`: equals 2010
- `age`: equals individuals' birth year

```{r}

# Add AGE and YEAR variables
bes10 <- bes10 %>%
  mutate(year = 2010,
         age = birthyr)

```


## Append and Merge

**12.** Append `bes10` to `bes05`. On doing so, you may wish to save the appended dataframe to your computer as a means of backing up your work.

```{r}

# Append
bes0510 <- bind_rows(bes05, bes10)

```

**13.** Merge the appended dataframe above and the data on casualties by UK region (`ukregion_cas`). The aim is to produce a dataframe that matches each individual in the BES data to the number of casualties in their UK region for 2005 and 2010. (Hint: You should merge using two "key" variables.)

```{r}

# Merge
bes0510casmerge <- bes0510 %>%
  left_join(casualties, by = c("region","year"))

```

**14.** Merge the merged dataframe above and the data on UK district democgraphics (`districtdata`). The aim is to produce a dataframe that matches each individual to their district demographics for 2005 and 2010. (Hint: You should merge using two "key" variables.)

```{r}

# Merge
bes_final_data <- bes0510casmerge %>%
  left_join(districts, by = c("region","year"))

```

 ## Manipulate (Again)

**15.** Create the following five variables. Add them to the dataframe that you created above. Here are the variable definitions:

- `white`: dummy variable that equals 1 if an individual is white and 0 otherwise.
- `female`: dummy variable that equals 1 if an individual is female and otherwise.
- `low_attention`: dummy variable that equals 1 if an individuals' political attention is less than 4 and 0 otherwise.
- `married`: dummy variable that equals 1 if an individual is married and 0 otherwise.
- `partstrength`: dummy variable that equals 1 if and individuals "very strongly" identifies with a political party and 0 otherwise.

```{r}

# Create WHITE, FEMALE, LOW_ATTENTION, MARRIAGE, and PARTSTRENGTH variables.
bes_final_data <- bes_final_data %>%
  mutate(white = ifelse(race == 1, 1, 0),
         female = ifelse(gender == 2, 1, 0),
         low_attention = ifelse(attention < 4, 1, 0),
         married = ifelse(marital_status == 1, 1, 0),
         partstrength = ifelse(party_strength == 1, 1, 0))

```

## Explore

**16.** Generate summary statistics for the following variables. For each, generate the minimum, maximum, median, mean, standard deviation, and number of observations. See if you can use `dplyr`'s summary functionality to create and print a new dataframe of summary statistics.

- `likelyvote`
- `region_cas`
- `low_attention`
- `female`
- `married`
- `income`
- `education`
- `age`
- `white`
- `partstrength`
- `perception_economy`
- `pmtherm`
- `pmwar`
- `unemploy_rate`
- `income_pc`
- `pct_white`

```{r}

my_vars <- c("likelyvote", "region_cas", "low_attention", "female", "married", "income", 
             "education", "age", "white", "partstrength", "perception_economy", "pmtherm",
             "pmwar", "unemploy_rate", "income_pc", "pct_white")

summary_df <- bes_final_data %>%
  mutate(across(everything(), as.numeric)) %>%
  summarise(across(everything(), list(
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE),
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    n = ~sum(!is.na(.))
  ), .names = "{.col}__{.fn}"))

summary_df <- summary_df %>%
  pivot_longer(everything(), names_to = c("Variable", "Statistic"), names_sep = "__") %>%
  pivot_wider(names_from = "Statistic", values_from = "value")


summary_df %>%
  knitr::kable(
    format = "latex",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    digits = 2,
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )

```

**17.** Generate seperate visulizations of the distributions of `region_cas`, `low_attention`, and `likelyvote`. Be sure to select visualizations that are appropriate for variables' type. Use `ggplot2` syntax. 

```{r}

p1 <- ggplot(bes_final_data, aes(region_cas)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = "Distribution of War Casualties (Iraq and Afghanistan) by UK Region\n")
p1

```

```{r}

bes_final_data <- bes_final_data %>%
  mutate(low_attention = as.factor(low_attention))

p2 <- ggplot(bes_final_data, aes(as.factor(low_attention))) +
  geom_bar() +
  scale_x_discrete(labels = c("High", "Low")) +
  theme_bw() +
  labs(title = "Level of Attention to Politics among UK Respondents\n",
       x = "\nLevel of Attention to Politics",
       y = "Number of Repondents\n") 
p2

```

```{r}

p3 <- ggplot(bes_final_data, aes(as.numeric(likelyvote))) +
  geom_bar() +
  scale_x_continuous(breaks = seq(0,10,1)) +
  theme_bw() +
  labs(title = "Likelihood of Voting among Repspondents\n",
       x = "\nLikelihood of Voting",
       y = "Number of Repondents\n") 
p3

```
